u---
title: "Team Project 5"
output: html_notebook
---

## Load data
```{r}
# importing files
require(readr)
# data processing
require(dplyr)
require(tidyr)
require(lubridate) # for date processing

# Map making
library(picante)
library(sf)
library(tmap)

# Visualization
library(ggplot2)
# library(ggtree)

# Qbio
## Phylogenies
library(phyloregion)
library(ape)
require('vegan')

# Beta diversity
library(pvclust)
``` 

```{r}
mydata <- read_csv(file = "DeamDataInitial.csv")
```

## Keep desired columns
```{r}
df1 <- subset(mydata, select = c(family, genus, specificEpithet, recordedBy, eventDate, country, stateProvince, county, decimalLatitude, decimalLongitude, geodeticDatum, coordinateUncertaintyInMeters))
```


## Combine genus and specificEpithet and then remove original columns
```{r}
df1$speciesName <- paste(df1$genus,df1$specificEpithet)
df2 <- subset(df1, select = c(family, speciesName, recordedBy, eventDate, country, stateProvince, county, decimalLatitude, decimalLongitude))
```

## Remove data with no GPS coordinates or dates
```{r}
df3 <- df2 %>% drop_na()
```

## Remove data that are not identified to species (denoted by "NA") and remove hybrids (denoted by "√ó")?
```{r}
library(tidyverse)
df4 <- df3 %>% filter_all(all_vars(!grepl("NA", .)))
df5 <- df4 %>% filter_all(all_vars(!grepl("ó", .)))
```

## Correct county name spelling
```{r}
## Upload county name list
ctydata <- read.csv("ctydata.csv", header = TRUE, fileEncoding="UTF-8-BOM")
# Test for wrong county spellings
x <- df5$county %in% ctydata$county
y <- data.frame(x)
z <- data.frame(which(grepl("FALSE", y$x)))
## If z contains data fix
## Correct county name spelling
df5$county[df5$county == "knox"] <- "Knox"
df5$county[df5$county == "gibson"] <- "Gibson"
df5$county[df5$county == "cass"] <- "Cass"
df5$county[df5$county == "pike"] <- "Pike"
df5$county[df5$county == "orange"] <- "Orange"
df5$county[df5$county == "grant"] <- "Grant"
df5$county[df5$county == "wayne"] <- "Wayne"
df5$county[df5$county == "jasper"] <- "Jasper"
df5$county[df5$county == "ELkhart"] <- "Elkhart"
df5$county[df5$county == "crawford"] <- "Crawford"
# Test for wrong county spellings
x <- df5$county %in% ctydata$county
y <- data.frame(x)
z <- data.frame(which(grepl("FALSE", y$x)))
## If z contains data fix
```

## Correct Indiana name spelling
```{r}
# Check for records not Indiana
table(df5[,5])
# Correct indiana to Indiana
df5$stateProvince[df5$stateProvince == "indiana"] <- "Indiana"
# Check again
table(df5[,5])
```

## Correct species names against official list at: https://midwestherbaria.org/portal/checklists/checklist.php?clid=3510&pid=124&dynclid=0. Note that 4 species not on that list are included here because they are exotic species validly documented in the state.
```{r}
## Upload county name list
indspecies <- read.csv("indspecies.csv", header = TRUE, fileEncoding="UTF-8")
# Test for wrong species spellings
x <- df5$speciesName %in% indspecies$ScientificName
y <- data.frame(x)
z <- data.frame(which(grepl("FALSE", y$x)))
## Fix data
df5$speciesName[df5$speciesName == "Andropogon gerardii"] <- "Andropogon gerardi"
df5$speciesName[df5$speciesName == "Adicea deamii"] <- "Pilea pumila"
df5$speciesName[df5$speciesName == "Baptisia lactea"] <- "Baptisia alba"
df5$speciesName[df5$speciesName == "Lacinaria scariosa"] <- "Liatris scariosa"
df5$speciesName[df5$speciesName == "Carex striatula"] <- "Carex laxiflora"
df5$speciesName[df5$speciesName == "Viburnum pubescens"] <- "Viburnum dentatum"
df5$speciesName[df5$speciesName == "Celtis pumila"] <- "Celtis occidentalis"
df5$speciesName[df5$speciesName == "Vernonia altissima"] <- "Vernonia gigantea"
df5$speciesName <- as.character(df5$speciesName)
df5$speciesName[grepl("Malus io", df5$speciesName)] <- "Malus ioensis"
df5$speciesName <- as.factor(df5$speciesName)
df5$speciesName[df5$speciesName == "Botrychium obliquum"] <- "Sceptridium dissectum"
df5$speciesName[df5$speciesName == "Arabis viridis"] <- "Borodinia missouriensis"
df5$speciesName <- as.character(df5$speciesName)
df5$speciesName[grepl("tes engelmannii", df5$speciesName)] <- "Isoetes engelmannii"
df5$speciesName <- as.factor(df5$speciesName)
df5$speciesName[df5$speciesName == "Juncus macer"] <- "Juncus anthelatus"
df5$speciesName[df5$speciesName == "Solidago deamii"] <- "Solidago simplex"
df5$speciesName[df5$speciesName == "Rosa deamii"] <- "Rosa carolina"
df5$speciesName[df5$speciesName == "Panicum deamii"] <- "Dichanthelium ovale"
df5$speciesName <- as.character(df5$speciesName)
df5$speciesName[df5$speciesName == "Piptatherum pungens"] <- "Piptatheropsis pungens"
df5$speciesName[df5$speciesName == "Helianthus rigidus"] <- "Helianthus pauciflorus"
df5$speciesName[df5$speciesName == "Malus angustifolia"] <- "Malus ioensis"
# Remove remaining hybrids
df6 <- df5[!df5$speciesName == "Lacinaria deamii", ]
df7 <- df6[!df6$speciesName == "Crataegus incaedua", ]
# Check species again
x <- df7$speciesName %in% indspecies$ScientificName
y <- data.frame(x)
z <- data.frame(which(grepl("FALSE", y$x)))
```

## Export df7 to csv as filtered data
```{r}
write.csv(df7, "Deam_Collection_Data_Filtered_3-27-2021.csv")
```

## Create site-species matrix for the complete dataset, treating counties as sites
```{r}
# Subset data to only include species name and county
df8 <- subset(df7, select = c(county, speciesName))
# Sum number of occurrences of each species in each county

df9 <- df8 %>% count(county, speciesName)
# Create site x species matrix by reshaping data
compdatasitesp <- pivot_wider(df9, names_from = speciesName, values_from = n,)
# Replace NA's with 0
compdatasitesp[is.na(compdatasitesp)] <- 0

# convert back to data frame and name counties for sites. 
compdatasitesp <- as.data.frame(compdatasitesp)
rownames(compdatasitesp) <- compdatasitesp[,1]
compdatasitesp <- compdatasitesp[,-1]
```

## SPATIAL PHYLOGENETIC ANALYSIS

## Start spatial phylogenetic analysis of woody plants

## Import woody tree species phylogeny and visualize
```{r}
# Import phlogenetic tree
# Data from: 10.1111/jbi.12171
woodytreesphylo <- read.tree(file = "Woody_Trees_Phylogeny.txt")
# Plot phylogeny
plot(woodytreesphylo, show.tip.label=FALSE)
```

## Subset Deam data by woody tree species present in phylogeny
```{r}
# Replace "_" symbol with " " in phylogeny label data
woodytreesphylo$tip.label <- gsub("_", " ", woodytreesphylo$tip.label, fixed=TRUE)
# Establish dataframe with tree species names
woodytreesp <- data.frame(woodytreesphylo$tip.label)
colnames(woodytreesp)[1] <- "speciesName"
# Subset Deam data using species names in phylogeny
woodytreedeam <- df7 %>%
      filter(df7$speciesName %in% woodytreesp$speciesName)
```

## Calculate number of species in Indiana and number of occurrences for each species
```{r}
# Number of occurrences of each species
speciescount <- setNames(as.data.frame(table(woodytreedeam$speciesName)), c("speciesName", "n"))
# Number of occurences per county
countycount <- setNames(as.data.frame(table(woodytreedeam$county)), c("county", "n"))
countycount$county <- as.character(countycount$county)
countycount$county[countycount$county == "La Porte"] <- "LaPorte"
```

# Create site species matrix for woody tree species in Indiana
```{r}
# Subset data to only include species name and county
woodytreedeamssm1 <- subset(woodytreedeam, select = c(county, speciesName))
woodytreedeamssm2 <- woodytreedeamssm1 %>% count(county, speciesName)
# Create site x species matrix by reshaping data
woodytreedeamssm3 <- pivot_wider(woodytreedeamssm2, names_from = speciesName, values_from = n,)
# Replace NA's with 0
woodytreedeamssm3[is.na(woodytreedeamssm3)] <- 0
# convert back to data frame and name counties for sites. 
woodytreedeamssm3 <- as.data.frame(woodytreedeamssm3)
rownames(woodytreedeamssm3) <- woodytreedeamssm3[,1]
woodytreedeamssm3 <- woodytreedeamssm3[,-1]
```

# Calculate Faith's phylogenetic diversity per Indiana county
```{r}
# install and load package
#install.packages("picante")

# Calculate Faith's phylogenetic diversity
fpdwoody <- pd(woodytreedeamssm3, woodytreesphylo, include.root=TRUE)
# Set row header to new column
fpdwoody$county <- rownames(fpdwoody)
# Change La Porte to "LaPorte"
fpdwoody$county[fpdwoody$county == "La Porte"] <- "LaPorte"
```

# Upload Indiana county maps
```{r warning=FALSE}
# upload shapefile with indiana counties
indcounties1 <- st_read("Indiana_Counties.shp")
# subset indiana counties
indcounties2 <- subset(indcounties1, select = c(NAME, geometry))
# change crs to wgs84
indcounties3 <- st_transform(indcounties2, 4326)
# rename "NAME" column to "county"
colnames(indcounties3)[1] <- "county"
# insert county centroid point
indcounties4 <- indcounties3
indcounties4$centroid <- st_centroid(indcounties4$geometry)
indcounties4$xy <- st_coordinates(indcounties4$centroid)
```

# Plot heat map of Faith's phylogenetic diversity for woody trees of Indiana on per county basis
```{r}
# Merge data
map <- merge(indcounties3,
                     fpdwoody,
                     by.x = "county",
                     by.y = "county")
# Plot
tm_shape(map, projection="+init=epsg:4326") +
  tm_polygons("PD", border.col = "grey30", title="") +
tm_layout(title="Faith's Phylogenetic Diversity", 
          title.size = 1, 
          title.position = c("center", "top"), 
          legend.text.size=.6, bg.color = "light blue", inner.margins=c(.1,.1,.2,.4))
tm_legend(position = c("right", "bottom"), 
	frame = TRUE,
	bg.color="lightblue")
```

## Calculate and plot alpha diversity of woody tree species in Indiana
```{r}
# define species richness function with site species matrix as input
S.obs <- function(x = ""){
  rowSums(x > 0) * 1
}
# calculate site species richness (county) using site species matrix of Indiana woody tree species
obssprichwt <- data.frame(S.obs(woodytreedeamssm3))
colnames(obssprichwt)[1] <- "speciesRichness"
# Set row header to new column
obssprichwt$county <- rownames(obssprichwt)
# Change La Porte to "LaPorte"
obssprichwt$county[obssprichwt$county == "La Porte"] <- "LaPorte"
# Plot species richness as heat map per county
# Merge data
map1 <- merge(indcounties3,
                     obssprichwt,
                     by.x = "county",
                     by.y = "county")
# Plot
tm_shape(map1, projection="+init=epsg:4326") +
  tm_polygons("speciesRichness", border.col = "grey30", title="") +
tm_layout(title="Observed Species Richness", 
          title.size = 1, 
          title.position = c("center", "top"), 
          legend.text.size=.6, bg.color = "light blue", inner.margins=c(.1,.1,.2,.4))
tm_legend(position = c("right", "bottom"), 
	frame = TRUE,
	bg.color="lightblue")
```

## Test for significance of woody tree alpha diversity in southern section of state
```{r}
# import county data
ctydatalat <- read.csv("IndianaCountiesSN.csv", header = TRUE, fileEncoding="UTF-8-BOM")
# merge with species richness data
ctyspeciesrich <- merge(ctydatalat,
                     obssprichwt,
                     by.x = "county",
                     by.y = "county")
# perform t-test
t.test(speciesRichness ~ latitude, alternative = "less", data = ctyspeciesrich)
```

## Test for correlation between alpha diversity and phylogenetic diversity
```{r}
# summarize data
summary(fpdwoody)
# check for normality
hist(fpdwoody$PD)
hist(fpdwoody$SR)
# take log of data
fpdwoody$logpd <- log(fpdwoody$PD)
fpdwoody$logsr <- log(fpdwoody$SR)
# perform linear regression
pd.sr.lm <- lm(logsr ~ logpd, data = fpdwoody)
summary(pd.sr.lm)
# visualize linear regression
alphaphylo.graph <- ggplot(fpdwoody, aes(x=logpd, y=logsr))+
                     geom_point() + geom_smooth(method="lm", col="black") + labs(title = "Species Richness vs. Phlyogenetic Diversity",
      x = "Faith's Phylogenetic Diversity (log)",
      y = "Species Richness (log)")
alphaphylo.graph
```

## Test for unequal scaling between alpha and phylogenetic diversity across the state
```{r}
# Calculate PD/SR ratio
fpdwoody$pdsrratio <- fpdwoody$PD / fpdwoody$SR
# import county data
ctydatascaling <- read.csv("IndianaCountiesScaling.csv", header = TRUE, fileEncoding="UTF-8-BOM")
# merge with species richness data
ctyspeciesscale <- merge(ctydatascaling,
                     fpdwoody,
                     by.x = "county",
                     by.y = "county")
ctyspeciesscale$colors <- as.factor(ctyspeciesscale$latitude)
# run anova
anovapdsrratio <- aov(pdsrratio ~ latitude, data = ctyspeciesscale)
summary(anovapdsrratio)
# visualize data via boxplot
boxplot(pdsrratio ~ latitude, data = ctyspeciesscale, main = "PD:AD Ratio across Three Regions",
   xlab = "Indiana Regions", ylab = "PD:AD")
# Plot of data
alphaphylo.graph <- ggplot(ctyspeciesscale, aes(x=logpd, y=logsr,col=colors))+
                     geom_point() + geom_smooth(method="lm", col="black") + labs(title = "Species Richness vs. Phlyogenetic Diversity",
      x = "Faith's Phylogenetic Diversity (log)",
      y = "Species Richness (log)")
alphaphylo.graph
```


## Calculate and plot beta diversity of woody tree species in Indiana
```{r}
# create matrix with counties as columns
transspsite <- pivot_wider(woodytreedeamssm2, names_from = county, values_from = n,)
# Replace NA's with 0
transspsite[is.na(transspsite)] <- 0
# convert back to data frame and name counties for sites. 
transspsite <- as.data.frame(transspsite)
rownames(transspsite) <- transspsite[,1]
transspsite <- transspsite[,-1]
# perform cluster analysis
#woodytree.ward <- pvclust(transspsite, method.hclust = "ward.D2", iseed = 65, nboot = 10000)
# plot cluster
#par(mar = c(1, 5, 2, 2) + 0.1)
#plot(woodytree.ward, main = "Woody Tree Species in Indiana: Ward's Clustering", ylab = "Squared Bray-Curtis Distance", cex = 0.6, col.pv = c(au = 3, bp = FALSE, edge = FALSE))
#function from: http://www.jafy.eu/posts/2019/06/pvclust-nodevalues-in-ggtree.md/
as.phylo.pvclust.node.attributes <- function(x, attribute) {
  N <- dim(x$merge)[1]
  edge <- matrix(0L, 2*N, 2)
  edge.length <- numeric(2*N)
  ## `node' gives the number of the node for the i-th row of x$merge
  node <- integer(N)
  node[N] <- N + 2L
  node.attributes <- rep(NA, N)
  cur.nod <- N + 3L
  j <- 1L
  for (i in N:1) {
    edge[j:(j + 1), 1] <- node[i]
    for (l in 1:2) {
      k <- j + l - 1L
      y <- x$merge[i, l]
      if (y > 0) {
        edge[k, 2] <- node[y] <- cur.nod
        cur.nod <- cur.nod + 1L
        edge.length[k] <- x$height[i] - x$height[y]
        node.attributes[edge[k, 1] - (N + 1)] <- attribute[i]
      } else {
        edge[k, 2] <- -y
        edge.length[k] <- x$height[i]
        node.attributes[edge[k, 1] -  (N + 1)] <- attribute[i]
      }
    }
    j <- j + 2L
  }
  
  if (is.null(x$labels))
    x$labels <- as.character(1:(N + 1))
  
  ## MODIFICATION: clean up node.attributes so they are in same format in 
  ## pvclust plots
  node.attributes <- as.character(round(node.attributes * 100, 0))
  node.attributes[1] <- NA
  
  obj <- list(edge = edge, edge.length = edge.length / 2,
              tip.label = x$labels, Nnode = N, node.label = node.attributes)
  class(obj) <- "phylo"
  stats::reorder(obj)
}
# perform function
hclust_boot_phylo <- as.phylo.pvclust.node.attributes(woodytree.ward$hclust, woodytree.ward$edges$bp)
# plot
ggtree(hclust_boot_phylo) +
  geom_tippoint(aes(subset = !is.na(label) & as.numeric(label) > 50 & as.numeric(label) <= 70, label = label),fill = 'white',pch=23,size = 3) +
  geom_tippoint(aes(subset = !is.na(as.numeric(label)) & as.numeric(label) > 70, label = label),fill = 'white',pch=21,size = 3) +
  geom_tippoint(aes(subset = !is.na(as.numeric(label)) & as.numeric(label) > 80, label = label),fill = 'gray',pch=21,size = 3) +
  geom_tippoint(aes(subset = !is.na(as.numeric(label)) & as.numeric(label) > 95, label = label),fill = 'black',pch=21,size = 3)
```

# Test for ecoregion clustering
```{r}
# upload county ecoregion dataframe
ctydataecoreg <- read.csv("IndianaCountiesEcoregionsIV.csv", header = FALSE, fileEncoding="UTF-8-BOM")
rownames(ctydataecoreg) <- ctydataecoreg[,1]
ctydataecoreg <- ctydataecoreg[,-1]
as.factor(ctydataecoreg)
# run PERMANOVA with adonis function
adonis(woodytreedeamssm3 ~ ctydataecoreg, method = "bray", permutations = 999)
# upload county ecoregion dataframe
ctydataecoregiii <- read.csv("IndianaCountiesEcoregionsIII.csv", header = FALSE, fileEncoding="UTF-8-BOM")
rownames(ctydataecoregiii) <- ctydataecoregiii[,1]
ctydataecoregiii <- ctydataecoregiii[,-1]
as.factor(ctydataecoregiii)
# run PERMANOVA with adonis function
adonis(woodytreedeamssm3 ~ ctydataecoregiii, method = "bray", permutations = 999)
```


# Create heat map of occurrences of woody tree species in Indiana
```{r}
# Merge data
map3 <- merge(indcounties3,
                     countycount,
                     by.x = "county",
                     by.y = "county")
# Plot
tm_shape(map3, projection="+init=epsg:4326") +
  tm_polygons("n", border.col = "grey30", title="") +
tm_layout(title="Number of Occurrences of Woody Tree Species", 
          title.size = 1, 
          title.position = c("center", "top"), 
          legend.text.size=.6, bg.color = "light blue", inner.margins=c(.1,.1,.2,.4))
tm_legend(position = c("right", "bottom"), 
	frame = TRUE,
	bg.color="lightblue")
```




# Census Data Associations with Species

```{r}
S.obs <- function(x=''){
  rowSums(x>0)*1
}
``` 

What Census should we use?

```{r}
a <- parse_date_time(as.vector(df7$eventDate),orders = c("mdy","ymd"))
# Items with invalid Dates:
# df7[which(is.na(a)),]
mean(a,na.rm=TRUE)
median(a,na.rm=TRUE)
``` 


```{r}
# US population Census 1920
USCensus <- as.data.frame(read.csv("data/IndianaCensus.csv",header=TRUE))
rownames(USCensus) <- USCensus[,1]
USCensus <- USCensus[,-1]
# Census of Agriculture of 1920
AgCensus <- as.data.frame(read.csv("data/IndianaAgCensus1920.csv",header=TRUE))
rownames(AgCensus) <- AgCensus[,1]
AgCensus <- AgCensus[,-1]
# Glaciation
Glacier <- as.data.frame(read.csv("data/Glaciation.csv",header=TRUE))
#Glacier$Glaciation<- as.factor(Glacier$Glaciation)
# Ecoregions
EcoIII <- as.data.frame(read.csv("IndianaCountiesEcoregionsIII.csv",header=FALSE,col.names = c("county","EcoIII")))
#EcoIII$EcoIII <- as.factor(EcoIII$EcoIII)
EcoIV <- as.data.frame(read.csv("IndianaCountiesEcoregionsIV.csv",header=FALSE,col.names = c("county","EcoIV")))
#EcoIV$EcoIV <- as.factor(EcoIV$EcoIV)
``` 

```{r}
popData <- cbind(USCensus[-1,],AgCensus[-1,],EcoIII[-1],Glacier[-1],EcoIV[-1],ctyspeciesscale$latitude)
samplingEffort <- as.data.frame(rowSums(compdatasitesp))
colnames(samplingEffort) <- c("sampEffort")
popData <-cbind(popData,samplingEffort)

popData$pcImproved1920 <- popData$ImprovedFarmAcres1920/popData$LandArea1920
popData$popDensity1920 <- popData$X1920/popData$LandArea1920


woodytreedeamssm3.db <- vegdist(woodytreedeamssm3,method='bray',binary = TRUE,upper=TRUE,diag=TRUE)

woodytreedeamssm3.dbrda <- dbrda(woodytreedeamssm3.db ~.,as.data.frame(popData))
ordiplot(woodytreedeamssm3.dbrda)
``` 
Cool, from this it is clear that the census population levels are minimally correlated, while the land area of counties and improved acres are more correlated. 

```{r}
woodytreedeamssm3.dbrda.mod0 <-dbrda(woodytreedeamssm3.db~1,as.data.frame(popData))
# ordiplot(woodytreedeamssm3.dbrda.mod0)
``` 

```{r}
woodytreedeamssm3.dbrda.mod1 <- dbrda(woodytreedeamssm3.db~., as.data.frame(popData))
# ordiplot(woodytreedeamssm3.dbrda.mod1)
``` 

```{r}
woodytreedeamssm3.dbrda <- ordiR2step(woodytreedeamssm3.dbrda.mod0,woodytreedeamssm3.dbrda.mod1,perm.max=600)

woodytreedeamssm3.dbrda$anova
ordiplot(woodytreedeamssm3.dbrda)
``` 
```{r}
scores(woodytreedeamssm3.dbrda,display="bp")
```
OOF only 4% of variation explained by these the agricultural information we inputted, which probably isn't that bad given that we are talking about species diversity across an entire state. Let's see what happens with the information given by the spatial correlations.
 

```{r}
rs <- rowSums(woodytreedeamssm3)/sum(woodytreedeamssm3)
woodytreedeamssm3.pcnmw <- pcnm(dist(indcounties4$xy),w=rs,dist.ret = TRUE)

woodytreedeamssm3.space <-as.data.frame(scores(woodytreedeamssm3.pcnmw))
woodytreedeamssm3.pcnm.mod0 <-dbrda(woodytreedeamssm3.db ~1,woodytreedeamssm3.space)
woodytreedeamssm3.pcnm.mod1 <-dbrda(woodytreedeamssm3.db ~.,woodytreedeamssm3.space)
step.pcnm <- ordiR2step(woodytreedeamssm3.pcnm.mod0, woodytreedeamssm3.pcnm.mod1,perm.max=500)
``` 

```{r}
step.pcnm$anova

``` 

We'll this is interesting. It looks like space doesn't actually influence the presence of woody species that much. In the case of using BC, the results are that the null model is chosen, in the case of Sorensen's it is that only 1.4% of variation is accounted for spatially, with the model of PCNM37. 

```{r}
space.mod <- model.matrix(~PCNM37, as.data.frame(woodytreedeamssm3.space))[,-1]
farm.mod <- model.matrix(~pcImproved1920 + ImprovedFarmAcres1920, as.data.frame(popData))[,-1]
``` 

```{r}
woodytreedeamssm3.total.farm <- dbrda(woodytreedeamssm3.db ~farm.mod)
``` 

```{r}
woodytreedeamssm3.varpart <- varpart(woodytreedeamssm3.db,farm.mod,space.mod)
woodytreedeamssm3.varpart
plot(woodytreedeamssm3.varpart)
``` 

### Repeat analysis for all species

```{r}
compdatasitesp.db <- vegdist(compdatasitesp,method='bray',binary = TRUE,upper=TRUE,diag=TRUE)

compdatasitesp.dbrda <- dbrda(compdatasitesp.db ~.,as.data.frame(popData))
ordiplot(compdatasitesp.dbrda)
``` 
Cool, from this it is clear that the census population levels are minimally correlated, while the land area of counties and improved acres are more correlated. 

```{r}
compdatasitesp.dbrda.mod0 <-dbrda(compdatasitesp.db~1,as.data.frame(popData))
# ordiplot(compdatasitesp.dbrda.mod0)
``` 

```{r}
compdatasitesp.dbrda.mod1 <- dbrda(compdatasitesp.db~., as.data.frame(popData))
# ordiplot(compdatasitesp.dbrda.mod1)
``` 

```{r}
compdatasitesp.dbrda <- ordiR2step(compdatasitesp.dbrda.mod0,compdatasitesp.dbrda.mod1,perm.max=200)

compdatasitesp.dbrda$anova
ordiplot(compdatasitesp.dbrda)
``` 

This is actually better with a 10% from these two variables 
 

```{r}
rs <- rowSums(compdatasitesp)/sum(compdatasitesp)
compdatasitesp.pcnmw <- pcnm(dist(indcounties4$xy),w=rs,dist.ret = TRUE)

compdatasitesp.space <-as.data.frame(scores(compdatasitesp.pcnmw))
compdatasitesp.pcnm.mod0 <-dbrda(compdatasitesp.db ~1,compdatasitesp.space)
compdatasitesp.pcnm.mod1 <-dbrda(compdatasitesp.db ~.,compdatasitesp.space)
step.pcnm <- ordiR2step(compdatasitesp.pcnm.mod0, compdatasitesp.pcnm.mod1,perm.max=500)
``` 


```{r}
step.pcnm$anova
``` 

#need to change 
```{r}
space.mod <- model.matrix(~PCNM53, as.data.frame(compdatasitesp.space))[,-1]
farm.mod <- model.matrix(~pcImproved1920 + ImprovedFarmAcres1920, as.data.frame(popData))[,-1]
``` 

```{r}
compdatasitesp.total.farm <- dbrda(compdatasitesp.db ~farm.mod)
``` 

```{r}
compdatasitesp.varpart <- varpart(compdatasitesp.db,farm.mod,space.mod)
compdatasitesp.varpart
plot(compdatasitesp.varpart)
``` 

```{r}
popData$county <- rownames(popData)
popData["La Porte","county"]="LaPorte"
map <- merge(indcounties3,
                     popData,
                     by.x = "county",
                     by.y = "county")
# Plot
tm_shape(map, projection="+init=epsg:4326") +
  tm_polygons("pcImproved1920", border.col = "grey30", title="") +
tm_layout(title="Percent Improved Agricultural Land", 
          title.size = 1, 
          title.position = c("center", "top"), 
          legend.text.size=.6, bg.color = "light blue", inner.margins=c(.1,.1,.2,.4))
tm_legend(position = c("right", "bottom"), 
	frame = TRUE,
	bg.color="lightblue")
```

```{r}
# Plot
tm_shape(map, projection="+init=epsg:4326") +
  tm_polygons("sampEffort", border.col = "grey30", title="") +
tm_layout(title="Sampling Effort for all species", 
          title.size = 1, 
          title.position = c("center", "top"), 
          legend.text.size=.6, bg.color = "light blue", inner.margins=c(.1,.1,.2,.4))
tm_legend(position = c("right", "bottom"), 
	frame = TRUE,
	bg.color="lightblue")
``` 

